Coursera Reproducible Research: Peer Assesment 1
===============================================

  
## Loading and preprocessing the data

Load the data:
```{r Loading the data, echo=TRUE}
activity.data <- read.csv("data/activity.csv")          
```
Change `date` variable from factor to date
```{r Preprocessing the data, echo=TRUE}
activity.data$date <- as.Date(activity.data$date)          
```

  
## What is mean total number of steps taken per day?

Make a histogram of the total number of steps taken each day:
```{r Histogram of total number of steps, echo=TRUE}
hist(activity.data$steps, main="Total no. of steps taken each day",
     xlab="Steps taken")
```

Calculate the mean and median of total number of steps taken per day:
```{r Mean and median, echo=TRUE}
mean(activity.data$steps, na.rm=TRUE)
median(activity.data$steps, na.rm=TRUE)
```

  
## What is the average daily activity pattern?

Calculate average number of steps taken by interval:
```{r Average steps by interval, echo=TRUE}
steps.ave <- tapply(activity.data$steps, activity.data$interval, mean, 
                    na.rm=TRUE)
steps.ave <- as.numeric(steps.ave)
```

Make a time series plot:
```{r Time series plot, echo=TRUE}
plot(steps.ave ~ activity.data$interval[1:288], type="l", 
     main="Average daily activity pattern", xlab="Interval", 
     ylab="Average no. of steps")
```

Interval which conatains the maximum numbers of steps is:
```{r Interval with max no. of steps, echo=TRUE}
activity.data$interval[which.max(steps.ave)]
```

  
## Imputing missing values

Calculate the total number of missing values in the dataset:
```{r No. of missing values, echo=TRUE}
sum(is.na(activity.data))
```

Replace missing values in the dataset with the mean for respective 5-minute interval:
```{r Replace missing values, echo=TRUE}
ave.steps.int <- rep(steps.ave, 61)
new.activity.data <- cbind(activity.data, ave.steps.int)

for (i in 1:nrow(new.activity.data)) {
        
        if (is.na(new.activity.data[i, 1])) {
                
                new.activity.data[i, 1] <- new.activity.data[i, 4]
        
        }        

}
```

Create a new dataset that is equal to the original dataset but with the missing data filled in: 
```{r New dataset, echo=TRUE}
new.activity.data <- new.activity.data[ ,-4]
head(activity.data)
head(new.activity.data)
```

Make a histogram of the total number of steps taken each day:
```{r Histogram of total number of steps (NAs filled in), echo=TRUE}
hist(new.activity.data$steps, main="Histogram of Steps taken each day",
     xlab="Steps taken")
```

Calculate the mean and median of total number of steps taken per day:
```{r Mean and median of new dataset, echo=TRUE}
mean(new.activity.data$steps)
median(new.activity.data$steps)
```

Do these values differ from the estimates from the first part of the assignment? **No**  
What is the impact of imputing missing data on the estimates of the total daily number of steps? **There is none**

  
## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day:
```{r New factor variable, echo=TRUE}
for (i in 1:nrow(new.activity.data)) {
        
        if (weekdays(new.activity.data[i, 2]) == "Saturday" 
            | weekdays(new.activity.data[i, 2]) == "Sunday") {
                
                new.activity.data[i, 4] <- "weekend"
        
        } else {
                
                new.activity.data[i, 4] <- "weekday"
        
        }

}

names(new.activity.data)[4] <- "day.type"
new.activity.data$day.type <- as.factor(new.activity.data$day.type)
```

Transform dataset for easier ploting:
```{r Transform dataset for easier ploting, echo=TRUE}
# Split dataset by day.type (weekday and weekend datasets)
split.activity <- split(new.activity.data, new.activity.data$day.type)
weekday <- split.activity[[1]]
weekend <- split.activity[[2]]

# Calculate average number of steps taken by interval for weekdays
steps.weekday <- tapply(weekday$steps, weekday$interval, mean, 
                        na.rm=TRUE)  

# Calculate average number of steps taken by interval for weekends
steps.weekend <- tapply(weekend$steps, weekend$interval, mean, 
                        na.rm=TRUE)

# Create new dataset for xyplot() function
steps.weekday <- as.numeric(steps.weekday)
steps.weekend <- as.numeric(steps.weekend)

steps.weekday.int <- data.frame(steps=steps.weekday, 
                                interval=new.activity.data$interval[1:288], 
                                day.type=rep("weekday", 288))
steps.weekend.int <- data.frame(steps=steps.weekend, 
                                interval=new.activity.data$interval[1:288], 
                                day.type=rep("weekend", 288))

activity.day.type <- rbind(steps.weekday.int, steps.weekend.int)
```

Make a panel plot:
```{r Panel plot, echo=TRUE}
library(lattice)
xyplot(activity.day.type$steps ~ activity.day.type$interval | activity.day.type$day.type, type="l", xlab="Interval", ylab="Number of steps", layout=c(1, 2))
```